\vspace{-0.5cm}The X-shooter GTO sample of GRB afterglow and host Galaxy spectra

### 2016-12-07 document/XSGRB_sample.tex
Change to A&A and HI

\bibliographystyle{aa}
\bibpunct{(}{)}{;}{a}{}{,}    %% natbib cite format used by A&A and ApJ

\newcommand{\farc}{\hbox{$.\!\!^{\prime\prime}$}} 
\newcommand{\ergA}{$\rm{erg\,cm^{-2}\,s^{-1}\,\AA^{-1}}$} 
\newcommand{\erg}{$\rm{erg\,cm^{-2}\,s^{-1}}$} 
\newcommand{\kms}{$\rm{km\,s^{-1}}\,$}

\newcommand{\griz}{$g' r' i' z'$}
\newcommand{\JHK}{$JHK_{\rm{s}}$}
\newcommand{\gK}{$g' r' i' z' JHK_{\rm{s}}$}
\newcommand{\Msun}{$M_\odot$}
\newcommand{\lya}{Ly$\alpha$} 
\newcommand{\lyb}{Ly$\beta$} 
\newcommand{\lyg}{Ly$\gamma$} 
\newcommand{\hb}{H$\beta$} 
\newcommand{\ha}{H$\alpha$} 
\newcommand{\hg}{H$\gamma$} 
\newcommand{\hi}{\ion{H}{1}} 
\newcommand{\hei}{[\ion{He}{1}]} 
\newcommand{\oi}{[\ion{O}{1}]} 
\newcommand{\sii}{[\ion{S}{2}]} 
\newcommand{\siii}{[\ion{S}{3}]} 
\newcommand{\oii}{[\ion{O}{2}]} 
\newcommand{\oiii}{[\ion{O}{3}]}
\newcommand{\nii}{[\ion{N}{2}]} 
\newcommand{\nv}{[\ion{N}{5}]} 
\newcommand{\neiii}{[\ion{Ne}{3}]} 
\newcommand{\NIii}{[\ion{Ni}{2}]} 
\newcommand{\feii}{\ion{Fe}{2}} 
\newcommand{\civ}{\ion{C}{4}} 
\newcommand{\cii}{\ion{C}{2}}
\newcommand{\mgi}{\ion{Mg}{1}} 
\newcommand{\mgii}{\ion{Mg}{2}} 
\newcommand{\ali}{\ion{Al}{1}} 
\newcommand{\alii}{\ion{Al}{2}} 
\newcommand{\aliii}{\ion{Al}{3}} 
\newcommand{\SIii}{\ion{Si}{2}} 
\newcommand{\SIiv}{\ion{Si}{4}} 
\newcommand{\znii}{\ion{Zn}{2}} 
\newcommand{\crii}{\ion{Cr}{2}} 
\newcommand{\mnii}{\ion{Mn}{2}} 
\newcommand{\tiii}{\ion{Ti}{2}} 
\newcommand{\caii}{\ion{Ca}{2}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria and observations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800 \AA~in
a single exposure, by splitting the light in three separate spectroscopic arms
using dichroics.
This way each arm work as a separate instrument, each functioning as its own
echelle spectrograph.
The ultraviolet blue(UVB) arm covering 3000 - 5500 \AA, the visual(VIS) arm
covering 5500 - 10200 \AA, and the near-infrared()NIR) arm covering 10200 -
24800 \AA~ with the possibility of applying a k-band
blocking filter cutting the coverage of the NIR arm at 10200 - 21000 \AA.

@@

For the large majority of the bursts we have observed with a slit width of
1\farc0, 0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively, which
puts a lower limit on the delivered resolution of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}.
The slit width also sets the width of the atmospheric sky lines and determines
the amount of light lost due to the wavelength-dependent seeing PSF extending
outside the coverage of the slit, where the width of sky-lines is always set by
the slit width whereas both the delivered resolution and the slitloss changes
for the better as the seeing PSF drops below the slit width.
For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9
observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit,
meaning that for almost all observations a slitloss correction is required. We
describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements of velocity widths, a precise instrumental resolution
is required and this becomes better when the delivered seeing is better than the
slit width. We discuss this in Sect. \ref{resolution}.

\todo{ADC discussion}

@@

For all earth-based telescopes the light of interest has to pass through Earths
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be calculated in a multitude of
ways and we employ a prioritized list of methods here, depending on the
availability of the different method. Since the observation are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the optimal correction.

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
\citep{Smette2015, Kausch2015}, which can then be to correct for the absorption.
The requirement here is that the afterglow continuum spectrum has a
signal-to-noise higher than a value of 10.

@@

%\input{tables/burst_overview.tex}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{X-ray column densities vs. optically derived ones}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
	\centerline{\includegraphics[width=10cm]{figures/OAXR_NH_sphere.pdf}}
	\caption{Optically derived HI column densities vs. X-ray derived ones taken from \citet{Buchner2016}. The colors of the points reflect the redshift of the burst. Circles mark the column densities published in \citet{Fynbo2009} and squares mark the ones derived as part of the sample presented here. Overplot in blue is the 1:1 correspondence, expected if the conditions local to the GRBs are similar to the local ISM abundances. If the metallicity is lower at the GRB site, the point move upwards in the plot. }
	\label{fig:OAXR}
\end{figure}

@@

\subsection{GRB100425A (z=1.1755)}
The data presented here also formed the basis of GCN \#
10684\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10684.gcn3}} and is used in
\citet{Skuladottir2010}, but not published elsewhere. Observations started 4
hours after the \textit{Swift} trigger, totaling 2.4 ks. Absorption features
from \mgii~and \feii~in the afterglow continuum are detected at $z=1.1755$.

>>>>

\bibliographystyle{mnras}
\bibpunct{(}{)}{;}{a}{}{,}    %% natbib cite format used by A&A and ApJ

% Manually specified definitions
\input{definitions.tex}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and observations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800 \AA~in
a single exposure, by splitting the light in three separate spectroscopic arms
using dichroics.
This way each arm work as a separate instrument, each functioning as its own
echelle spectrograph.
The ultraviolet blue(UVB) arm covering 3000 - 5500 \AA, the visual(VIS) arm
covering 5500 - 10200 \AA, and the near-infrared()NIR) arm covering 10200 -
24800 \AA~ with the possibility of applying a k-band
blocking filter cutting the coverage of the NIR arm at 10200 - 21000 \AA.
\\

@@

For the large majority of the bursts we have observed with a slit width of
1\farc0, 0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively, which
puts a lower limit on the delivered resolution of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}.
The slit width also sets the width of the atmospheric sky lines and determines
the amount of light lost due to the wavelength-dependent seeing PSF extending
outside the coverage of the slit, where the width of sky-lines is always set by
the slit width whereas both the delivered resolution and the slitloss changes
for the better as the seeing PSF drops below the slit width.
For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9
observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit,
meaning that for almost all observations a slitloss correction is required. We
describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements of velocity widths, a precise instrumental resolution
is required and this becomes better when the delivered seeing is better than the
slit width. We discuss this in Sect. \ref{resolution}.
\\
\todo{ADC discussion}

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex}

@@

For all earth-based telescopes, the light of interest has to pass through Earths
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be calculated in a multitude of
ways and we employ a prioritized list of methods here, depending on the
availability of the different method. Since the observation are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the optimal correction.

The highest priority method is using the GRB afterglow continuum itself, where
the atmospheric conditions have directly been imprinted on the spectrum. The
telluric features can directly be fit with an atmospheric model
\citep{Smette2015, Kausch2015}, which can then be used to correct for the absorption. The accuracy of the correction depends on the 
The requirement here is that the afterglow continuum spectrum has a
signal-to-noise higher than a value of 10.

@@

@@

Because the locations of GRBs are associated with the most intensely starforming
regions \citep{Hogg1999, Bloom2002, Fruchter2006}, the GRB afterglow light has
to propagate through the large amounts of hydrogen fueling the star formation.
Because a significant fraction of the hydrogen has not been ionized yet, the
optical depth at the wavelength of \lya~is very high, saturating at the line
center. This causes a strong absorption system to appear in the afterglow
continuum. For bursts with $z \gtrsim 1.7$, the position of \lya~moves into the
spectroscopic coverage of X-shooter, meaning that we are able to detect this
absorption trough due to \lya.

Due of the stochastic nature of the \lya-forest, the blue wing of the Lyman
alpha absorption line is randomly superposed with forest systems, making it
notoriously difficult to model. Similarly so, the red wing has got the ISM
signatures imprinted on it, especially strong absorption due to \SIii~which can
exhibit significant velocity structure. Along with instrumental effects, the
generative model for the data that we would use in a likelihood-based analysis
would be almost boundless in complexity, making formal fitting of the column
densities unfeasible.

Using an analytic approximation to an absorption line profile and it's
dependence on the column density \citep{TepperGarcia2006}, we overplot a
synthetic absorption line with a specified column density on our observed
spectrum. By tuning the value of the hydrogen column density until the synthetic
absorption line matches the spectrum, we can thereby infer the actual column
density of the GRB sight line in a manual way. In a similar fashion, the errors
on the hydrogen column can be inferred. We show the results of this procedure
for all bursts where possible in Fig. \ref{fig:HI1} and the inferred hydrogen
column densities in Tab. \ref{tab:HI}

\input{tables/HI_columns.tex}

\begin{figure*}
	\centering
	\includegraphics[page=1, width=16cm]{figures/HI_measurements.pdf}
	\caption{ Measurements of the hydrogen column-densities for all bursts with a clear Lyman alpha absorption system. In solid black is shown the spectrum with black dotted giving the corresponding 1-$\sigma$ error. Black dashed shows zero flux density. The solid red line is the absorption of column density equal to the value presented in Tab. \ref{HI_tab} with the 1-$\sigma$ interval shown with dashed lines.}
	\label{fig:HI1}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=2, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI2}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=3, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI3}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=4, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI4}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=5, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI5}
\end{figure*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{X-ray column densities vs. optically derived ones}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%
%\begin{figure}
%	\centerline{\includegraphics[width=10cm]{figures/OAXR_NH_sphere.pdf}}
%	\caption{Optically derived HI column densities vs. X-ray derived ones taken from \citet{Buchner2016}. The colors of the points reflect the redshift of the burst. Circles mark the column densities published in \citet{Fynbo2009} and squares mark the ones derived as part of the sample presented here. Overplot in blue is the 1:1 correspondence, expected if the conditions local to the GRBs are similar to the local ISM abundances. If the metallicity is lower at the GRB site, the point move upwards in the plot. }
%	\label{fig:OAXR}
%\end{figure}

@@

\subsection{GRB100425A (z=1.1755)}
The data presented here also formed the basis of GCN \#
10684\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10684.gcn3}} and is used in
Skuladottir (2010), but not published elsewhere. Observations started 4
hours after the \textit{Swift} trigger, totaling 2.4 ks. Absorption features
from \mgii~and \feii~in the afterglow continuum are detected at $z=1.1755$.
